Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-24 paper.tex
Standardise on VCF Zarr
Closes #66

% Additional packages used for code formatting
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}

@@

\textbf{Results:}
We present the vcfzarr specification, a columnar encoding of the 
VCF data model using Zarr which makes retrieving subsets of the 
data much more efficient. Zarr is a cloud-native format for storing 
multi-dimensional numerical data, widely used in scientific computing.
We show how this format is competitive with specialised methods for 
storing genotype data in terms of compression ratios, and also performs
well on compressing the high-entropy quality-control fields that 
tend to dominate. We show that performing calculations on the genotype
matrix is much more efficient than standard
approaches based on VCF, and is comparable with state-of-the-art methods
using compiled languages. We demonstrate 
the vcfzarr format (and the vcf2zarr conversion utility) 
on the Genomics England dataset of XX,XXX,XXX samples, showing
an overall reduction in storage of XX, and a reduction in processing 
time of common queries of up to XX fold.

\textbf{Conclusions:}
Large row-encoded VCF files are a major bottleneck for current research, and 
processing these files incurs a substantial economic cost.
The vcfzarr specification, building on widely-used, open-source technologies
has the potential to greatly reduce these costs,
and may enable a diverse ecosystem of next-generation tools for analysing 
genetic variation data.
\end{abstract}

@@

\section{Background}
[Four paragraphs. (1) What is VCF, and the good things it has done. 
(2) Explosion in dataset size. Summary of where we're at in terms of biobanks
etc. (3) We have a problem. Existing solutions like TileDB etc
are either proprietary, or tightly coupled to one specific software stack
(Hail, e.g.).
(4) We present the vcfzarr spec and the vcf2zarr conversion utility.]

@@

\begin{figure}
\resizebox{225pt}{!}{\input{diagrams/data-model.tex}}
\caption{Columnar storage of VCF data in chunked, compressed form using Zarr.
[TODO more detail, including brief overview of how data maps from 
VCF to vcfzarr in the examples]
\label{fig-data-model}}
\end{figure}

We propose the \texttt{vcfzarr} specification which maps the 
VCF data model into a columnar layout using the
widely-used Zarr format (Fig~\ref{fig-data-model}). 
Briefly, Zarr is a format 
% FIXME two different uses of $n$ here in a paragraph
for storing $n$-dimensional arrays of a fixed datatype, stored as 
compressed chunks. In the \texttt{vcfzarr} specification, 
each field in a VCF is mapped to an array, which is compressed 
and stored individually allowing for efficient retrieval. 
In particular, call-level data is stored as $m \times n$ arrays
(for $m$ sites and $n$ samples), allowing for efficient 
retrieval of subsets of those fields along both the 
variants and samples axis.
Note that this approach to storing higher dimensional data
sets Zarr apart from more general columnar formats 
such as Parquet [TODO more details and citations?].
See the Methods for more detail on Zarr and the \texttt{vcfzarr}
specification.

@@

\subsection{Case study: Genomics England 100,000 genomes}
In this section we demonstrate the utility of \texttt{vcfzarr} on a large human dataset
and the scalability of the \texttt{vcf2zarr} conversion utility.
Genomics England’s multi-sample VCF dataset (aggV2) is an 
aggregate of 78,195 gVCFs from rare disease and cancer participants 
recruited as part of the 100,000 Genomes Project~\cite{turnbull2018100}. 
The dataset comprises approximately 722 million annotated single-nucleotide 
variants (SNVs) and small indels split into 1,371 roughly equal chunks and 
totalling 165.3 TiB of VCF data after \texttt{bgzip} compression. 
The dataset is used for a variety of research purposes, ranging from 
genome-wide association studies (GWAS)~\cite{kousathanas2022whole} and 
imputation~\cite{shi2023genomics} to 
simple queries involving single gene 
regions~\cite{leggatt2023genotype,lam2023repeat}.

@@

\begin{table}
\caption{Summary for a selection of the largest vcfzarr columns produced for 
Genomics England aggV2 VCFs on chromosome 20 using \texttt{vcf2zarr}
default settings. Each field is stored independently 
as a Zarr array with the given type (sufficient to represent all values in the
data). We show the total storage consumed (reported via \texttt{du}) in 
power-of-two units, and the compression ratio achieved on that array.
We also show the percentage of the overall storage that each array consumes.
\label{tab-genomics-england-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
call\_AD &  int16 & 179.7G & 26 & 24.0\\
call\_GQ &  int16 & 171.8G & 13 & 23.0 \\
call\_DP &  int16 & 141.8G & 16 & 18.9 \\
call\_DPF& int16  & 115.1G & 20 & 15.3\\
call\_FT &  string & 58.5G & 160 & 7.8 \\
call\_PL &  int16 & 51.4G & 140 & 6.9 \\
call\_GQX &  int16 & 12.1G & 190 & 1.6 \\
call\_genotype & int8 & 6.1G & 380 & 0.8 \\
call\_genotype\_mask & bool & 3.7G  & 630 & 0.5\\
call\_genotype\_phased & bool & 692.5M  & 1700 & 0.1 \\
call\_PS  & int8  & 102.2M & 12000 & 0.05 \\
variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
variant\_allele & string & 21.5M & 11 \\
variant\_position & int32 & 12.7M & 4.7 \\
variant\_ABratio & float32 & 9.6M & 6.3 \\
variant\_AN & int32 & 9.6M & 6.3 \\
variant\_filter & bool & 1.9M & 490 \\
\bottomrule
\end{tabular}
\end{table}

@@

Nonetheless, at least in the single-shared-copy model currently
used by most large-scale variation datasets, Zarr provides
an excellent basis for building a new generation of tools 
that compute \emph{directly} on the data, efficiently
accessing data by variant and by sample.
[Mention ``Cloud native data repositories''~\cite{abernathey2021cloud}]
The vcfzarr specification and scalable vcf2zarr conversion utility
that we have provided here are the starting points of such 
an ecosystem, but much more would be needed to make a viable 
alternative for everyday bioinformatics workflows. However, a few key
pieces of infrastructure would provide a starting point, and would 
be sufficient to greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England.
Firstly, a \texttt{vcfzarrtools} command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) would immediately speed up 
limited region and field specific queries by orders of magnitude,
and would provide compatibility with existing workflows.
Developing such a utility would be a non-trivial but routine
engineering task.

>>>>

@@

\textbf{Results:}
We present the VCF Zarr specification, a columnar encoding of the 
VCF data model using Zarr which makes retrieving subsets of the 
data much more efficient. Zarr is a cloud-native format for storing 
multi-dimensional numerical data, widely used in scientific computing.
We show how this format is competitive with specialised methods for 
storing genotype data in terms of compression ratios, and also performs
well on compressing the high-entropy quality-control fields that 
tend to dominate. We show that performing calculations on the genotype
matrix is much more efficient than standard
approaches based on VCF, and is comparable with state-of-the-art methods
using compiled languages. We demonstrate 
the VCF Zarr format (and the vcf2zarr conversion utility) 
on the Genomics England dataset of XX,XXX,XXX samples, showing
an overall reduction in storage of XX, and a reduction in processing 
time of common queries of up to XX fold.

\textbf{Conclusions:}
Large row-encoded VCF files are a major bottleneck for current research, and 
processing these files incurs a substantial economic cost.
The VCF Zarr specification, building on widely-used, open-source technologies
has the potential to greatly reduce these costs,
and may enable a diverse ecosystem of next-generation tools for analysing 
genetic variation data.
\end{abstract}

@@

\section{Background}
[Four paragraphs. (1) What is VCF, and the good things it has done. 
(2) Explosion in dataset size. Summary of where we're at in terms of biobanks
etc. (3) We have a problem. Existing solutions like TileDB etc
are either proprietary, or tightly coupled to one specific software stack
(Hail, e.g.).
(4) We present the VCF Zarr spec and the vcf2zarr conversion utility.]

@@

\begin{figure}
\resizebox{225pt}{!}{\input{diagrams/data-model.tex}}
\caption{Columnar storage of VCF data in chunked, compressed form using Zarr.
[TODO more detail, including brief overview of how data maps from 
VCF to VCF Zarr in the examples]
\label{fig-data-model}}
\end{figure}

We propose the VCF Zarr specification which maps the 
VCF data model into a columnar layout using the
widely-used Zarr format (Fig~\ref{fig-data-model}). 
Briefly, Zarr is a format 
% FIXME two different uses of $n$ here in a paragraph
for storing $n$-dimensional arrays of a fixed datatype, stored as 
compressed chunks. In the VCF Zarr specification, 
each field in a VCF is mapped to an array, which is compressed 
and stored individually allowing for efficient retrieval. 
In particular, call-level data is stored as $m \times n$ arrays
(for $m$ sites and $n$ samples), allowing for efficient 
retrieval of subsets of those fields along both the 
variants and samples axis.
Note that this approach to storing higher dimensional data
sets Zarr apart from more general columnar formats 
such as Parquet [TODO more details and citations?].
See the Methods for more detail on Zarr and the VCF Zarr
specification.

@@

\subsection{Case study: Genomics England 100,000 genomes}
In this section we demonstrate the utility of VCF Zarr on a large human dataset
and the scalability of the \texttt{vcf2zarr} conversion utility.
Genomics England’s multi-sample VCF dataset (aggV2) is an 
aggregate of 78,195 gVCFs from rare disease and cancer participants 
recruited as part of the 100,000 Genomes Project~\cite{turnbull2018100}. 
The dataset comprises approximately 722 million annotated single-nucleotide 
variants (SNVs) and small indels split into 1,371 roughly equal chunks and 
totalling 165.3 TiB of VCF data after \texttt{bgzip} compression. 
The dataset is used for a variety of research purposes, ranging from 
genome-wide association studies (GWAS)~\cite{kousathanas2022whole} and 
imputation~\cite{shi2023genomics} to 
simple queries involving single gene 
regions~\cite{leggatt2023genotype,lam2023repeat}.

@@

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced for 
Genomics England aggV2 VCFs on chromosome 20 using \texttt{vcf2zarr}
default settings. Each field is stored independently 
as a Zarr array with the given type (sufficient to represent all values in the
data). We show the total storage consumed (reported via \texttt{du}) in 
power-of-two units, and the compression ratio achieved on that array.
We also show the percentage of the overall storage that each array consumes.
\label{tab-genomics-england-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
call\_AD &  int16 & 179.7G & 26 & 24.0\\
call\_GQ &  int16 & 171.8G & 13 & 23.0 \\
call\_DP &  int16 & 141.8G & 16 & 18.9 \\
call\_DPF& int16  & 115.1G & 20 & 15.3\\
call\_FT &  string & 58.5G & 160 & 7.8 \\
call\_PL &  int16 & 51.4G & 140 & 6.9 \\
call\_GQX &  int16 & 12.1G & 190 & 1.6 \\
call\_genotype & int8 & 6.1G & 380 & 0.8 \\
call\_genotype\_mask & bool & 3.7G  & 630 & 0.5\\
call\_genotype\_phased & bool & 692.5M  & 1700 & 0.1 \\
call\_PS  & int8  & 102.2M & 12000 & 0.05 \\
variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
variant\_allele & string & 21.5M & 11 \\
variant\_position & int32 & 12.7M & 4.7 \\
variant\_ABratio & float32 & 9.6M & 6.3 \\
variant\_AN & int32 & 9.6M & 6.3 \\
variant\_filter & bool & 1.9M & 490 \\
\bottomrule
\end{tabular}
\end{table}

@@

Nonetheless, at least in the single-shared-copy model currently
used by most large-scale variation datasets, Zarr provides
an excellent basis for building a new generation of tools 
that compute \emph{directly} on the data, efficiently
accessing data by variant and by sample.
[Mention ``Cloud native data repositories''~\cite{abernathey2021cloud}]
The VCF Zarr specification and scalable vcf2zarr conversion utility
that we have provided here are the starting points of such 
an ecosystem, but much more would be needed to make a viable 
alternative for everyday bioinformatics workflows. However, a few key
pieces of infrastructure would provide a starting point, and would 
be sufficient to greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England.
Firstly, a \texttt{vcfzarrtools} command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) would immediately speed up 
limited region and field specific queries by orders of magnitude,
and would provide compatibility with existing workflows.
Developing such a utility would be a non-trivial but routine
engineering task.
